### 2017-01-31 Theory/Theory.tex
Fixing up theory :-)

\subsection{Fundamental particles and forces}
\label{sec:theory_sm_particles}
In the \ac{SM} there are twelve fundamental fermions, six quarks and
six leptons. The quarks interact via both strong nuclear force and the 
electroweak force, while
the leptons only interact via the electoweak force.
Each of the fermions has a corresponding antiparticle
with opposite quantum numbers. The quarks and leptons
are divided into three generations. The particles from different
generations have similar properties, apart from the mass which increases
for each generation. The quarks and leptons are summarised in 
table \ref{tab:theory_fermions}.

@@

The forces in the \ac{SM} are mediated by spin-1 bosons: The gluon (\Pgluon),
which carries the strong force, the photon (\Pphoton) as the
electromagnetic force carrier, and the $\PW^{\pm}$ and \PZ bosons, the
force carriers of the weak nuclear force.

@@

Particles that carry electric charge interact via the 
electromagnetic force, summarised in by the rules of \ac{QED}. 
The weak interaction, governed by weak isospin transformations,
only couples to the left-handed part of the fermion states. In 
general a fermion state can be decomposed into its left- and right-handed parts
\begin{equation}\label{eqn:leftright}
\chi=\psi_R+\psi_L,
\end{equation}
and only $\psi_L$ will interact via the weak force. Handedness and
helicity have the same eigenstates for massless particles, and as
such neutrinos, which are massless in the \ac{SM}, must all be left-handed.
The observation of neutrino flavour oscillations since the formulation of
the \ac{SM} implies that neutrinos do have a small but finite mass
and so there must be right-handed neutrinos in the \ac{SM}. This is
not discussed further here ADD REFERENCE.
The weak isospin acts on doublets of left-handed lepton pairs, while
the right-handed leptons transform as singlet states. This works similarly
for the quarks, with weak isospin acting three left-handed doublets and
six right-handed singlets. Note that the weak eigenstates of the quarks
differ from the physical states, which mix via the \ac{CKM}-matrix CITATION
in charged current interactions as
\begin{equation}\label{eqn:ckm_matrix}
\begin{pmatrix} d' \\
s' \\
b'  \end{pmatrix} = \begin{pmatrix} V_{ud} & V_{us} & V_{ub} \\
V_{cd} & V_{cs} & V_{cb} \\
V_{td} & V_{ts} & V_{tb} \end{pmatrix} \begin{pmatrix} d \\
s\\
b \end{pmatrix}.
\end{equation}

\subsection{The electroweak force}
\label{sec:theory_sm_ewk}
The unification of the weak and electromagnetic forces, introduced
in the 1960s by Glashow \cite{glashow-ewk}, Weinberg \cite{weinberg-ewk} and Salam \cite{salam-ewk},
is a key component of the \ac{SM}. The unification of these forces,
governed by an SU(2)$\times$U(1) symmetry, implies the presence of a charged
current. A quantum number, weak hypercharge $Y$, is introduced. Weak
hypercharge is related to weak isospin $I_{1,2,3}$ and electric charge as
\begin{equation}\label{eqn:hypercharge}
Q=I_3 + \frac{1}{2}Y.
\end{equation}
We then have the three weak isospin fields, $W_{\mu}^{i}$, and the
weak hypercharge field $Y_{\mu}$, which mix to form the physical \Pphoton,
\PZ and $\PW^{\pm}$ states as
\begin{align}\label{eqn:ewk_propagators}
\PW^{\pm} &= \frac{1}{\sqrt{2}}(W_{\mu}^1 \mp i W_{\mu}^2),\\
A_{\mu} &= W_{\mu}^3\sin{\theta_W} + Y_{\mu}\cos{\theta_W},\\
Z_{\mu} &= W_{\mu}^3\cos{\theta_W} - Y_{\mu}\sin{\theta_W}.
\end{align}
In these equations $A_{\mu}$ is the photon field and $Z_{\mu}$ the \PZ field.
The weak mixing angle $\theta_W$ is related to the weak (g) and electromagnetic (g')
coupling constants as
\begin{equation}\label{eqn:thetaw}
g\sin{\theta_W} = g'\cos{\theta_W}
\end{equation}

The gauge invariance of the Lagrangian would be broken
by the addition of mass terms. For the photon this is fine, as we know it
is massless, but it is known that the \PW and \PZ bosons, as well as fermions, 
are massive. This means electroweak symmetry must be broken. In the \ac{SM}
this is achieved via the Higgs mechanism.

\subsection{The Higgs mechanism}
\label{sec:theory_sm_higgsmech}
The masses of the vector bosons are generated via the
Higgs mechanism, proposed in the 1960s by Englert and Brout \cite{englertbrout}

>>>>

\subsection{Fundamental particles}
\label{sec:theory_sm_particles}
In the \ac{SM} there are twelve fundamental fermions, six quarks and
six leptons. The quarks interact via both strong nuclear force and the 
electroweak force, while
the leptons only interact via the electoweak force.
Each of the fermions has a corresponding antiparticle
with opposite quantum numbers. The quarks and leptons
are divided into three generations. The particles from different
generations have similar properties, apart from the mass which increases
for each generation. The quarks and leptons are summarised in 
table \ref{tab:theory_fermions}.

@@

\subsection{Fundamental forces}
\label{sec:theory_sm_forces}
%Symmetries of the Lagrangian describing a physical system lead
%to conserved quantities, as shown by Noether's theorem \cite{noethersthm}.
%This is and important concept for the description of particle interactions
%under different forces: invariance (symmetry) under a certain 
%transformation group is imposed on 
%the Lagrangians describing the fundamental forces. This is brought
%about as local gauge invariance, where the Lagrangians 
%are required to be invariant under a local phase transformation
%plus a gauge transformation. The local phase transformation
%might introduce additional derivatives of potentials, taken care of by gauge transf.
The forces in the \ac{SM} are mediated by spin-1 bosons: The gluon (\Pgluon),
which carries the strong force, the photon (\Pphoton) as the
electromagnetic force carrier, and the $\PW^{\pm}$ and \PZ bosons, the
force carriers of the weak nuclear force.

@@

The weak and electromagnetic forces were unified into the
electroweak force. The unification of these forces was introduced 
in the 1960s by Glashow \cite{glashow-ewk}, Weinberg \cite{weinberg-ewk} and Salam \cite{salam-ewk},
is a key component of the \ac{SM}. Electroweak theory,
governed by an SU(2)$\times$U(1) symmetry, implies the presence of weak
neutral currents.

%Particles that carry electric charge interact via the
%electromagnetic force, while the weak force works
%between particles carrying weak isospin.
The weak interaction, working on all fermions,
only couples to the left-handed part of fermion states. In 
general a fermion state $\chi$ can be decomposed into its left- and right-handed parts
\begin{equation}\label{eqn:leftright}
\chi=\psi_R+\psi_L,
\end{equation}
and only $\psi_L$ will interact via the weak force. For massless particles
the general state $\chi$ is not a mixture of left-handed and right-handed parts,
but is either completely left-handed or right-handed. As such neutrinos, which are
massless in the \ac{SM}, must all be left-handed. %CAREFUL: handedness =/= helicity
The observation of neutrino flavour oscillations since the formulation of
the \ac{SM} implies that neutrinos do have a small but finite mass
and so there must be right-handed neutrinos in the \ac{SM}. This will
not be discussed here, for a more detailed account see reference \cite{pdg-2014}.
%The weak isospin acts on doublets of left-handed lepton pairs, while
%the right-handed leptons transform as singlet states. This works similarly
%for the quarks, with weak isospin acting three left-handed doublets and
Another feature of the weak force is that the weak eigenstates of the quarks
differ from the physical states, which mix via the \ac{CKM}-matrix \cite{ckm-matrix-cabibbo,ckm-matrix-KM}
in charged current interactions as
\begin{equation}\label{eqn:ckm_matrix}
\begin{pmatrix} d' \\
s' \\
b'  \end{pmatrix} = \begin{pmatrix} V_{ud} & V_{us} & V_{ub} \\
V_{cd} & V_{cs} & V_{cb} \\
V_{td} & V_{ts} & V_{tb} \end{pmatrix} \begin{pmatrix} d \\
s\\
b \end{pmatrix}.
\end{equation}

Going back to the symmetry of the unified electroweak forces,
The SU(2) part of the symmetry is built from the weak
isospin generators $I_{1,2,3} = \frac{1}{2}\sigma_{1,2,3}$,
where $\sigma_i$ are the Pauli spin matrices. The U(1)
symmetry comes from the weak hypercharge generator $Y$, which
is related to weak isospin and electric charge as $Q = I_3 + \frac{1}{2}Y$.

Associated gauge fields are the three weak fields $W_{\mu}^{i}$ and the
weak hypercharge field $B_{\mu}$. These four fields mix to form the physical \Pphoton,
\PZ and $\PW^{\pm}$ states as
\begin{equation}\label{eqn:ewk_propagators_W}
\PW^{\pm} &= \frac{1}{\sqrt{2}}(W_{\mu}^1 \mp i W_{\mu}^2),
\end{equation}
\begin{equation}\label{eqn:ewk_propagators_phot}
A_{\mu} &= W_{\mu}^3\sin{\theta_W} + Y_{\mu}\cos{\theta_W},
\end{equation}
\begin{equation}\label{eqn:ewk_propagators_Z}
Z_{\mu} &= W_{\mu}^3\cos{\theta_W} - Y_{\mu}\sin{\theta_W}.
\end{equation}
In these equations $A_{\mu}$ is the photon field and $Z_{\mu}$ the \PZ field.
The weak mixing angle $\theta_W$ is related to the weak (g) and electromagnetic (g')
coupling constants as
\begin{equation}\label{eqn:thetaw}
\begin{split}
&g\sin{\theta_W} = g'\cos{\theta_W} \Rightarrow \\
&\sin{\theta_W} = \frac{g'}{\sqrt{g^2+g'^2}} \text{ and } \cos{\theta_W} = \frac{g}{\sqrt{g^2+g'^2}}
\end{split}
\end{equation}

Neutral weak-current interactions were discovered in 1973 at the Gargamelle
bubble chamber experiment at \acs{CERN} \cite{gargamelle}. This discovery was 
followed 10 years later by the discovery of the $\PW^{\pm}$ \cite{UA1-1,UA2-1} and $\PZ$ \cite{UA1-2,UA2-2} bosons by the UA1
and UA2 collaborations at \acs{CERN}.

An issue with electroweak theory is that the gauge invariance of the Lagrangian would be
broken by the addition of mass terms for the gauge bosons. However, the \PW and \PZ bosons are known to be massive. 
Furthermore fermion mass terms would also not be gauge invariant, and it is also known
that the fermions are massive.
This means electroweak symmetry must be broken. In the \ac{SM}
this is achieved via the Higgs mechanism.

\subsection{The Higgs mechanism}
\label{sec:theory_sm_higgsmech}
The masses of the vector bosons are generated via the
Higgs mechanism, proposed in the 1960s by Englert and Brout \cite{englertbrout},
Higgs \cite{higgs-I,higgs-II,higgs-III} and Guralnik, Hagen and Kibble \cite{GHK}.
The idea is to add a field that is symmetric under gauge transformations, but
has a nonzero vacuum expectation value and thus breaks the symmetry spontaneously.
According to Goldstone's theorem \cite{goldstone-theorem,goldstone-theorem-2} a 
consequence of spontaneous symmetry breaking is the appearance of a 
massless scalar (Goldstone boson) for each broken generator.

The simplest addition to achieve this is a complex doublet,
\begin{equation}\label{eqn:phi_hfield}
\Phi = \begin{pmatrix} \phi^{+} \\
\phi^{0} \end{pmatrix}.
\end{equation}
This field has an associated Lagrangian
\begin{equation}\label{eqn:higgs_gen_lag}
\begin{split}
\mathcal{L}_{\text{Higgs}} = -(D_{\mu}\Phi)^{\dagger}(D^{\mu}\Phi) - V(\Phi^{\dagger}\Phi) \text{, where we choose, }\\
V = \lambda(\Phi^{\dagger}\Phi - \frac{\mu^2}{2\lambda})^2,
\end{split}
\end{equation}
Where $\lambda$ and $\mu^2$ are two real parameters, and $\lambda>0$ is needed for stability
of the vacuum. To get spontaneous symmetry breaking we require $\mu^2>0$.

Under a local $SU(2)\times U(1)$ transformation, with hypercharge Y=$\frac{1}{2}$
the resulting transformation of the field is,
\begin{equation}\label{eqn:hfield_trsf}
\begin{pmatrix} \phi^+ \\
\phi^0 \end{pmatrix} = \exp{\frac{i}{2}(\vec{\theta(x)}\cdot\vec{\sigma} + \rho(x))}
\begin{pmatrix} \phi^+ \\
\phi^0 \end{pmatrix}, 
\end{equation}
where $\vec{\sigma}$ are the Pauli spin matrices and $\vec{\theta(x)}$ and $\rho(x)$
are functions of the space-time coordinates. From such a transformation
the covariant derivative is required to be
\begin{equation}\label{eqn:H_cov_deriv}
D_{\mu} = \partial_{\mu} - \frac{i}{2}g\vec{W_{\mu}}\cdot\vec{\sigma} - \frac{i}{2}g'Y_{\mu}
\end{equation}
In the vacuum state we must have $V(\Phi^{\dagger}\Phi) =0$, which happens for nonzero values of $\Phi$,
and we can choose $\phi^+$ = 0 and $\phi^0$ real to get:
%\begin{equation}\label{eqn:vacuum_expt}
%\begin{split}
%&\Phi^{\dagger}\Phi - \frac{\mu^2}{2\lambda} = 0 \Rightarrow \\
%&\Phi^{\dagger}\Phi = \frac{\mu^2}{2\lambda} \Rightarrow\\
%&\phi^{+*}\phi^+ + \phi^{0*}\phi^0 = \frac{\mu^2}{2\lambda} \text{ and choosing } \phi^+ \text{ = 0 and } \phi^0 \text{ to be real,}\\
%&\phi^0 = \frac{v}{\sqrt{2}} \text{ where } v \text{ is real and} v^2=\frac{\mu^2}{\lambda}.
%\end{split}
%\end{equation}
\begin{equation}\label{eqn:field_vev}
\langle 0 | \Phi | 0 \rangle = \begin{pmatrix} 0 \\
\frac{v}{\sqrt{2}} \end{pmatrix},
\end{equation}
where $v$ is real and $v^2 = \frac{\mu^2}{\lambda}$.

Looking at fluctuations around the vacuum state,
\begin{equation}\label{eqn:fluctvac}
\begin{split}
&\delta\Phi = \frac{i}{2}(\vec{\theta}\cdot\vec{\sigma} + \rho)\Phi\\
&=\frac{i}{2}\begin{pmatrix} \theta_3 + \rho & \theta_1 - i\theta_2 \\
\theta_1+i\theta_2 & \rho - \theta_3 \end{pmatrix} \begin{pmatrix} 0\\
\frac{v}{\sqrt{2}} \end{pmatrix}.
\end{split}
\end{equation}
Requiring $\delta\Phi=0$, we have $\theta_1=\theta_2=0$ and $\rho=\theta_3$, 
and the only remaining generator that leaves the field 
invariant is $\frac{i}{2}\theta_3 + Y = I_3 +Y$ is the U(1) group. Denoting the 
remaining generators as $\vec{b}$, and parameterising the field as an expansion around the
vacuum state gives:
\begin{equation}\label{eqn:vev_expansion}
\Phi = \exp{\frac{i}{\sqrt{2}v}\vec{\theta}\cdot\vec{b}}\begin{pmatrix} 0 \\
\frac{1}{\sqrt{2}} (v+H) \end{pmatrix}.
\end{equation}
By choosing the appropriate gauge the Goldstone bosons $\vec{\theta}$
can be eliminated. They become the longitudinal degrees of freedom of the \PW and \PZ 
bosons. This leaves us with
\begin{equation}\label{eqn:vev_expansion_satisfied}
\Phi = \begin{pmatrix} 0 \\
\frac{1}{\sqrt{2}}(v+H) \end{pmatrix}
\end{equation}
Substituting this into the Lagrangian of equation \ref{eqn:higgs_gen_lag}, 
and using the definitions of $\PW_{\mu}^{\pm}$ and $\PZ_{\mu}$ from 
equations \ref{eqn:ewk_propagators}, we get
\begin{equation}\label{eqn:completed_lagrangian}
\mathcal{L}_{\text{Higgs}} = -\frac{1}{2}\partial_{\mu}H\partial^{\mu}H -\frac{g^2v^2}{4}\PW_{\mu}^- \PW^{\mu +} - \frac{g^2+g'^2}{8}v^2\PZ_{\mu}\PZ^{\mu}  - \mu^2H^2  + \text{terms of order 3 and up in the fields}.
\end{equation}
This gives us masses for the \PW bosons as $m_{\PW} = \frac{gv}{2}$ and a mass
for the \PZ boson as $m_{\PZ} = \frac{1}{2}v\sqrt{g^2+g'^2}$. The photon remains massles. 
The remaining mass term is for the field $H$, the
Higgs field, which has a mass $m_{H} = \sqrt{2\mu^2}$.
%Mass term : -1/2m_X^2. However for m_W note W_mu+Wmu- = 1/2(W_mu^1+iW_mu^2)(W_mu^1-iW_mu^2) = 
%1/2*(W_mu^1W_mu^1+W_mu^2W_mu^2). But also W+^2+W^-2 = W1^2+W2^2 so masses end up the same

Using the developments so far we can also add fermion masses
through Yukawa couplings between the fermion- and Higgs fields
\begin{equation}\label{eqn:yukawa_coupl}
\lambda_f(\bar{\psi_L}\Phi\psi_R + \bar{\psi_R}\Phi\psi_L),
\end{equation}
where $\lambda_f$ are coupling constants specific to each fermion, $\lambda_f \propto m_f$.
This means that heavier fermions have stronger couplings to the Higgs boson.
